---
title: "PHASE 2"
author: "Nimo Hassan, Alexander Wodarski, Breanna Ranglall, Tenzin Dhondup"
format: pdf
editor: visual
---

```{r}
#| warning: false
library(tidyverse)
library(readxl)
library(dplyr)
```

```{r}
# Read data from excel spreadsheet
dataf <- read_excel("OP-18_Waste_Minimization_and_Diversion_2024.xlsx")
pop_data <- read.csv("updated_school_population.csv")
```

```{r}
#| warning: false
dataframe <- dataf

# Converting 'chr' to 'numeric'

dataframe$'Materials recycled, performance year' <- as.numeric(dataframe$'Materials recycled, performance year')
dataframe$'Materials recycled, baseline year' <- as.numeric(dataframe$'Materials recycled, baseline year')
dataframe$'Materials composted, performance year' <- as.numeric(dataframe$'Materials composted, performance year')
dataframe$'Materials composted, baseline year' <- as.numeric(dataframe$'Materials composted, baseline year')
dataframe$'Materials donated or re-sold, performance year' <- as.numeric(dataframe$'Materials donated or re-sold, performance year')
dataframe$'Materials donated or re-sold, baseline year' <- as.numeric(dataframe$'Materials donated or re-sold, baseline year')
dataframe$'Materials disposed through post-recycling residual conversion, performance year' <- as.numeric(dataframe$'Materials disposed through post-recycling residual conversion, performance year')
dataframe$'Materials disposed through post-recycling residual conversion, baseline year' <- as.numeric(dataframe$'Materials disposed through post-recycling residual conversion, baseline year')
dataframe$'Materials disposed in a solid waste landfill or incinerator, performance year' <- as.numeric(dataframe$'Materials disposed in a solid waste landfill or incinerator, performance year')
dataframe$'Materials disposed in a solid waste landfill or incinerator, baseline year' <- as.numeric(dataframe$'Materials disposed in a solid waste landfill or incinerator, baseline year')
dataframe$'Total waste generated, performance year' <- as.numeric(dataframe$'Total waste generated, performance year')
dataframe$'Total waste generated, baseline year' <- as.numeric(dataframe$'Total waste generated, baseline year')
dataframe$'Start date, performance year or 3-year period' <- as.numeric(dataframe$'Start date, performance year or 3-year period')
dataframe$'End date, performance year or 3-year period' <- as.numeric(dataframe$'End date, performance year or 3-year period')
dataframe$'Start date, baseline year or 3-year period' <- as.numeric(dataframe$'Start date, baseline year or 3-year period')
dataframe$'End date, baseline year or 3-year period' <- as.numeric(dataframe$'End date, baseline year or 3-year period')
dataframe$'Number of students resident on-site, performance year' <- as.numeric(dataframe$'Number of students resident on-site, performance year')
dataframe$'Number of students resident on-site, baseline year' <- as.numeric(dataframe$'Number of students resident on-site, baseline year')
dataframe$'Number of employees resident on-site, performance year' <- as.numeric(dataframe$'Number of employees resident on-site, performance year')
dataframe$'Number of employees resident on-site, baseline year' <- as.numeric(dataframe$'Number of employees resident on-site, baseline year')
dataframe$'Number of other individuals resident on-site, performance year' <- as.numeric(dataframe$'Number of other individuals resident on-site, performance year')
dataframe$'Number of other individuals resident on-site, baseline year' <- as.numeric(dataframe$'Number of other individuals resident on-site, baseline year')
dataframe$'Total full-time equivalent student enrollment, performance year' <- as.numeric(dataframe$'Total full-time equivalent student enrollment, performance year')
dataframe$'Total full-time equivalent student enrollment, baseline year' <- as.numeric(dataframe$'Total full-time equivalent student enrollment, baseline year')
dataframe$'Full-time equivalent of employees, performance year' <- as.numeric(dataframe$'Full-time equivalent of employees, performance year')
dataframe$'Full-time equivalent of employees, baseline year' <- as.numeric(dataframe$'Full-time equivalent of employees, baseline year')
dataframe$'Full-time equivalent of students enrolled in exclusively in distance education, performance year' <- as.numeric(dataframe$'Full-time equivalent of students enrolled in exclusively in distance education, performance year')
dataframe$'Full-time equivalent of students enrolled exclusively in distance education, baseline year' <- as.numeric(dataframe$'Full-time equivalent of students enrolled exclusively in distance education, baseline year')
dataframe$'Weighted campus users, performance year' <- as.numeric(dataframe$'Weighted campus users, performance year')
dataframe$'Weighted campus users, baseline year' <- as.numeric(dataframe$'Weighted campus users, baseline year')
dataframe$'Total waste generated per weighted campus user, performance year' <- as.numeric(dataframe$'Total waste generated per weighted campus user, performance year')
dataframe$'Total waste generated per weighted campus user, baseline year' <- as.numeric(dataframe$'Total waste generated per weighted campus user, baseline year')
dataframe$'Weighted campus users, performance year' <- as.numeric(dataframe$'Weighted campus users, performance year')
dataframe$'Weighted campus users, baseline year' <- as.numeric(dataframe$'Weighted campus users, baseline year')
dataframe$'Percentage reduction in total waste generated per weighted campus user from baseline' <- as.numeric(dataframe$'Percentage reduction in total waste generated per weighted campus user from baseline')
dataframe$'Percentage of materials diverted from the landfill or incinerator by recycling, composting, ...' <- as.numeric(dataframe$'Percentage of materials diverted from the landfill or incinerator by recycling, composting, ...')
dataframe$'Percentage of materials diverted from the landfill or incinerator (including up to 10 percent ...' <- as.numeric(dataframe$'Percentage of materials diverted from the landfill or incinerator (including up to 10 percent ...')
# Replace the single NA with 0
# Replace NA values with 0 for the specified columns
dataframe$`Total waste generated, baseline year`[is.na(dataframe$`Total waste generated, baseline year`)] <- 0
dataframe$`Weighted campus users, baseline year`[is.na(dataframe$`Weighted campus users, baseline year`)] <- 0
dataframe$`Total waste generated, performance year`[is.na(dataframe$`Total waste generated, performance year`)] <- 0
dataframe$`Weighted campus users, performance year`[is.na(dataframe$`Weighted campus users, performance year`)] <- 0
dataframe$`Materials recycled, performance year`[is.na(dataframe$`Materials recycled, performance year`)] <- 0
dataframe$`Materials composted, performance year`[is.na(dataframe$`Materials composted, performance year`)] <- 0
dataframe$`Materials donated or re-sold, performance year`[is.na(dataframe$`Materials donated or re-sold, performance year`)] <- 0
dataframe$`Materials disposed in a solid waste landfill or incinerator, performance year`[is.na(dataframe$`Materials disposed in a solid waste landfill or incinerator, performance year`)] <- 0


```

```{r}
pop_data$Institution[pop_data$Institution == "University of Minnesota Twin Cities"] <- "University of Minnesota, Twin Cities"
pop_data$Institution[pop_data$Institution == "University of Minnesota Crookston"] <- "University of Minnesota, Crookston"
pop_data$Institution[pop_data$Institution == "University of Minnesota Morris"] <- "University of Minnesota, Morris"
```

```{r}
# Converting 'chr' to 'date'

dataframe$'Last Updated' <- as.Date(dataframe$'Last Updated')
dataframe$'Start date, performance year or 3-year period' <- as.Date(dataframe$'Start date, performance year or 3-year period', origin = "1899-12-30")
dataframe$'End date, performance year or 3-year period' <- as.Date(dataframe$'End date, performance year or 3-year period', origin = "1899-12-30")
dataframe$'Start date, baseline year or 3-year period' <- as.Date(dataframe$'Start date, baseline year or 3-year period', origin = "1899-12-30")
dataframe$'End date, baseline year or 3-year period' <- as.Date(dataframe$'End date, baseline year or 3-year period', origin = "1899-12-30")
pop_data$Undergraduate.Population <- as.numeric(pop_data$Undergraduate.Population)
```

```{r}
peer_schools <- c("Iowa State University", "University of Minnesota, Twin Cities", "University of St. Thomas", 
                  "Macalester College", "Gustavus Adolphus College", "University of Wisconsin-Madison", 
                  "Marquette University", "University of Wisconsin-Eau Claire", 
                  "University of Wisconsin-Green Bay", "University of Wisconsin-Oshkosh", 
                  "University of Wisconsin-Platteville", "University of Wisconsin-Stevens Point", 
                  "University of Wisconsin-River Falls", "University of Minnesota, Morris", 
                  "University of Minnesota, Crookston","Concordia University",
                  "Carleton College", "South Dakota State University", 
                  "University of South Dakota")
schools_to_exclude <- c("Marquette University", "University of Wisconsin-Green Bay", "University of Minnesota, Crookston")
peers <-  dataframe|>
  filter(Institution %in% peer_schools)|>
  left_join(pop_data,by = "Institution")|>
  filter(Undergraduate.Population<11000)|>
  filter(!Institution %in% schools_to_exclude)|>
  mutate(
  `Reduction in Total Waste per Person` = 5 * (
    (
      (`Total waste generated, baseline year` / `Weighted campus users, baseline year`) - 
      (`Total waste generated, performance year` / `Weighted campus users, performance year`)
    ) / 
    (`Total waste generated, baseline year` / `Weighted campus users, baseline year`)
  ),
  
  `Total waste per person` = 2.78 * (
    (
      0.46 - 
      (`Total waste generated, performance year` / `Weighted campus users, performance year`) / 0.46
    )
  ),
  
  `Waste diverted from the landfill or incinerator` = 3 * (
    (
      (
        `Materials recycled, performance year` + 
        `Materials composted, performance year` + 
        `Materials donated or re-sold, performance year` +
        (0.1 * (
          `Materials recycled, performance year` + 
          `Materials composted, performance year` + 
          `Materials donated or re-sold, performance year` + 
          `Materials disposed in a solid waste landfill or incinerator, performance year`
        ))
      ) / 
      (
        `Materials recycled, performance year` + 
        `Materials composted, performance year` + 
        `Materials donated or re-sold, performance year` + 
        `Materials disposed in a solid waste landfill or incinerator, performance year`
      )
    )
  )
)

           
# Summary statistics for all numeric variables
numeric_summary <- peers |>
  summarize(across(where(is.numeric), list(
    mean = ~ mean(.x, na.rm = TRUE),
    median = ~ median(.x, na.rm = TRUE),
    sd = ~ sd(.x, na.rm = TRUE),
    min = ~ min(.x, na.rm = TRUE),
    max = ~ max(.x, na.rm = TRUE)
  )))
View(peers)

```

While looking at the mn peer institutions we saw that there weren't many schools to compare to so we decided to gather schools from North Dakota, South Dakota, Iowa, and Wisconsin to have more schools to compare to. We made a new csv of those schools with their populations and we filtered it. From there we saw the big outliers so we filtered out any school with undergraduate populations over 11000. The last thing we did is filter out schools who didn't report data and couldn't have their scores calculated. Those are in the schools to exclude list. We decided to add the scores, we got the equations from the aaeshe stars website, and had to do a little work with the data to change the na variables to 0 so that each institution outputted a score.

```{r}
library(dplyr)


summarydata <- data.frame(
  
  Material = rep(c(
    "Materials recycled",
    "Materials composted",
    "Materials donated or re-sold",
    "Materials disposed through post-recycling residual conversion",
    "Materials disposed in a solid waste landfill or incinerator"
  ), each = 10),
  Year_Type = rep(c(
    "Performance Year", "Baseline Year"
  ), times = 5, each = 5),
  Metric = rep(c("Mean", "Median", "SD", "Min", "Max"), times = 10),
  Value = c(
    451.446, 229.74, 538.0773, 105, 1390.34,
    675.482, 525, 869.5525, 57, 2176.4,
    308.51, 67.4, 488.207, 53.05, 1176.1,
    215.175, 215.175, 208.7733, 67.55, 362.8,
    68.52, 6.6, 136.8534, 0, 312.84,
    107.52, 107.52, 63.32848, 62.74, 152.3,
    NA, NA, NA, Inf, -Inf,
    NA, NA, NA, Inf, -Inf,
    934.404, 358.09, 1558.256, 104, 3712.9,
    934.404, 358.09, 1558.256, 104, 3712.9
  )
  
)


organizedSummary <- summarydata|>
  pivot_wider(
    names_from = Metric,
    values_from = Value
  ) |>
  arrange(Material, Year_Type)


view(organizedSummary)

```

