---
title: "test_tidy"
author: "Alex Wodarski"
format: html
editor: visual
---

```{r}
#| warning: false
library(tidyverse)
library(readxl)
library(dplyr)
```

```{r}
# Read data from excel spreadsheet
data <- read_excel("OP-18_Waste_Minimization_and_Diversion_2024.xlsx")
```

```{r}
df <- data

# Converting 'chr' to 'numeric'
#| warning: false

df$'Materials recycled, performance year' <- as.numeric(df$'Materials recycled, performance year')
df$'Materials recycled, baseline year' <- as.numeric(df$'Materials recycled, baseline year')
df$'Materials composted, performance year' <- as.numeric(df$'Materials composted, performance year')
df$'Materials composted, baseline year' <- as.numeric(df$'Materials composted, baseline year')
df$'Materials donated or re-sold, performance year' <- as.numeric(df$'Materials donated or re-sold, performance year')
df$'Materials donated or re-sold, baseline year' <- as.numeric(df$'Materials donated or re-sold, baseline year')
df$'Materials disposed through post-recycling residual conversion, performance year' <- as.numeric(df$'Materials disposed through post-recycling residual conversion, performance year')
df$'Materials disposed through post-recycling residual conversion, baseline year' <- as.numeric(df$'Materials disposed through post-recycling residual conversion, baseline year')
df$'Materials disposed in a solid waste landfill or incinerator, performance year' <- as.numeric(df$'Materials disposed in a solid waste landfill or incinerator, performance year')
df$'Materials disposed in a solid waste landfill or incinerator, baseline year' <- as.numeric(df$'Materials disposed in a solid waste landfill or incinerator, baseline year')
df$'Total waste generated, performance year' <- as.numeric(df$'Total waste generated, performance year')
df$'Total waste generated, baseline year' <- as.numeric(df$'Total waste generated, baseline year')
df$'Start date, performance year or 3-year period' <- as.numeric(df$'Start date, performance year or 3-year period')
df$'End date, performance year or 3-year period' <- as.numeric(df$'End date, performance year or 3-year period')
df$'Start date, baseline year or 3-year period' <- as.numeric(df$'Start date, baseline year or 3-year period')
df$'End date, baseline year or 3-year period' <- as.numeric(df$'End date, baseline year or 3-year period')
df$'Number of students resident on-site, performance year' <- as.numeric(df$'Number of students resident on-site, performance year')
df$'Number of students resident on-site, baseline year' <- as.numeric(df$'Number of students resident on-site, baseline year')
df$'Number of employees resident on-site, performance year' <- as.numeric(df$'Number of employees resident on-site, performance year')
df$'Number of employees resident on-site, baseline year' <- as.numeric(df$'Number of employees resident on-site, baseline year')
df$'Number of other individuals resident on-site, performance year' <- as.numeric(df$'Number of other individuals resident on-site, performance year')
df$'Number of other individuals resident on-site, baseline year' <- as.numeric(df$'Number of other individuals resident on-site, baseline year')
df$'Total full-time equivalent student enrollment, performance year' <- as.numeric(df$'Total full-time equivalent student enrollment, performance year')
df$'Total full-time equivalent student enrollment, baseline year' <- as.numeric(df$'Total full-time equivalent student enrollment, baseline year')
df$'Full-time equivalent of employees, performance year' <- as.numeric(df$'Full-time equivalent of employees, performance year')
df$'Full-time equivalent of employees, baseline year' <- as.numeric(df$'Full-time equivalent of employees, baseline year')
df$'Full-time equivalent of students enrolled in exclusively in distance education, performance year' <- as.numeric(df$'Full-time equivalent of students enrolled in exclusively in distance education, performance year')
df$'Full-time equivalent of students enrolled exclusively in distance education, baseline year' <- as.numeric(df$'Full-time equivalent of students enrolled exclusively in distance education, baseline year')
df$'Weighted campus users, performance year' <- as.numeric(df$'Weighted campus users, performance year')
df$'Weighted campus users, baseline year' <- as.numeric(df$'Weighted campus users, baseline year')
df$'Total waste generated per weighted campus user, performance year' <- as.numeric(df$'Total waste generated per weighted campus user, performance year')
df$'Total waste generated per weighted campus user, baseline year' <- as.numeric(df$'Total waste generated per weighted campus user, baseline year')
df$'Weighted campus users, performance year' <- as.numeric(df$'Weighted campus users, performance year')
df$'Weighted campus users, baseline year' <- as.numeric(df$'Weighted campus users, baseline year')
df$'Percentage reduction in total waste generated per weighted campus user from baseline' <- as.numeric(df$'Percentage reduction in total waste generated per weighted campus user from baseline')
df$'Percentage of materials diverted from the landfill or incinerator by recycling, composting, ...' <- as.numeric(df$'Percentage of materials diverted from the landfill or incinerator by recycling, composting, ...')
df$'Percentage of materials diverted from the landfill or incinerator (including up to 10 percent ...' <- as.numeric(df$'Percentage of materials diverted from the landfill or incinerator (including up to 10 percent ...')
```

```{r}
# Converting 'chr' to 'date'

df$'Last Updated' <- as.Date(df$'Last Updated')
df$'Start date, performance year or 3-year period' <- as.Date(df$'Start date, performance year or 3-year period', origin = "1899-12-30")
df$'End date, performance year or 3-year period' <- as.Date(df$'End date, performance year or 3-year period', origin = "1899-12-30")
df$'Start date, baseline year or 3-year period' <- as.Date(df$'Start date, baseline year or 3-year period', origin = "1899-12-30")
df$'End date, baseline year or 3-year period' <- as.Date(df$'End date, baseline year or 3-year period', origin = "1899-12-30")
```

```{r}
peer_schools<-c("Bemidji State University","Carleton College","College of St. Benedict/St. John’s University", "Concordia Moorhead","Macalester College","Winona State University","University of Minnesota, Twin Cities","University of Minnesota, Morris","UMN – Duluth","Augsburg University","Concordia in St. Paul","Hamline University","St. Kate’s University","St. Olaf College","University of St. Thomas")
peers <-  df|>
  filter(Institution %in% peer_schools)
# Summary statistics for all numeric variables
numeric_summary <- peers |>
  summarize(across(where(is.numeric), list(
    mean = ~ mean(.x, na.rm = TRUE),
    median = ~ median(.x, na.rm = TRUE),
    sd = ~ sd(.x, na.rm = TRUE),
    min = ~ min(.x, na.rm = TRUE),
    max = ~ max(.x, na.rm = TRUE)
  )))
View(peers)
View(numeric_summary)
```

We can compare St. Thomas with all the summary statistics. One thing to note out of the peer institutions, is that the University of Minnesota Twin Cities skews the results as it has much larger numbers than all the other Institutions that we are comparing too. Because of this St. Thomas ends up being below the average for the statistics. The main assumption for the reason would be that the University of Minnesota Twin Cities is just a much larger school compared to the others on the list and therefore their numbers are much higher. St Thomas rank fluctuates as 2 or 3 for the variables behind the University of Minnesota, and they fluctuate with Carleton College whon has similar numbers for most. 