---
title: "Phase 3 Viz Breanna"
author: "Breanna Ranglall"
format: pdf
editor: visual
---

```{r}
#| warning: false 
library(tidyverse) 
library(readxl) 
library(dplyr)
```

```{r}
# Read data from excel spreadsheet 
dataf <- read_excel("OP-18_Waste_Minimization_and_Diversion_2024.xlsx") 
pop_data <- read.csv("updated_school_population.csv")
```

```{r}
#| warning: false
dataframe <- dataf

# Converting 'chr' to 'numeric'

dataframe$'Materials recycled, performance year' <- as.numeric(dataframe$'Materials recycled, performance year')
dataframe$'Materials recycled, baseline year' <- as.numeric(dataframe$'Materials recycled, baseline year')
dataframe$'Materials composted, performance year' <- as.numeric(dataframe$'Materials composted, performance year')
dataframe$'Materials composted, baseline year' <- as.numeric(dataframe$'Materials composted, baseline year')
dataframe$'Materials donated or re-sold, performance year' <- as.numeric(dataframe$'Materials donated or re-sold, performance year')
dataframe$'Materials donated or re-sold, baseline year' <- as.numeric(dataframe$'Materials donated or re-sold, baseline year')
dataframe$'Materials disposed through post-recycling residual conversion, performance year' <- as.numeric(dataframe$'Materials disposed through post-recycling residual conversion, performance year')
dataframe$'Materials disposed through post-recycling residual conversion, baseline year' <- as.numeric(dataframe$'Materials disposed through post-recycling residual conversion, baseline year')
dataframe$'Materials disposed in a solid waste landfill or incinerator, performance year' <- as.numeric(dataframe$'Materials disposed in a solid waste landfill or incinerator, performance year')
dataframe$'Materials disposed in a solid waste landfill or incinerator, baseline year' <- as.numeric(dataframe$'Materials disposed in a solid waste landfill or incinerator, baseline year')
dataframe$'Total waste generated, performance year' <- as.numeric(dataframe$'Total waste generated, performance year')
dataframe$'Total waste generated, baseline year' <- as.numeric(dataframe$'Total waste generated, baseline year')
dataframe$'Start date, performance year or 3-year period' <- as.numeric(dataframe$'Start date, performance year or 3-year period')
dataframe$'End date, performance year or 3-year period' <- as.numeric(dataframe$'End date, performance year or 3-year period')
dataframe$'Start date, baseline year or 3-year period' <- as.numeric(dataframe$'Start date, baseline year or 3-year period')
dataframe$'End date, baseline year or 3-year period' <- as.numeric(dataframe$'End date, baseline year or 3-year period')
dataframe$'Number of students resident on-site, performance year' <- as.numeric(dataframe$'Number of students resident on-site, performance year')
dataframe$'Number of students resident on-site, baseline year' <- as.numeric(dataframe$'Number of students resident on-site, baseline year')
dataframe$'Number of employees resident on-site, performance year' <- as.numeric(dataframe$'Number of employees resident on-site, performance year')
dataframe$'Number of employees resident on-site, baseline year' <- as.numeric(dataframe$'Number of employees resident on-site, baseline year')
dataframe$'Number of other individuals resident on-site, performance year' <- as.numeric(dataframe$'Number of other individuals resident on-site, performance year')
dataframe$'Number of other individuals resident on-site, baseline year' <- as.numeric(dataframe$'Number of other individuals resident on-site, baseline year')
dataframe$'Total full-time equivalent student enrollment, performance year' <- as.numeric(dataframe$'Total full-time equivalent student enrollment, performance year')
dataframe$'Total full-time equivalent student enrollment, baseline year' <- as.numeric(dataframe$'Total full-time equivalent student enrollment, baseline year')
dataframe$'Full-time equivalent of employees, performance year' <- as.numeric(dataframe$'Full-time equivalent of employees, performance year')
dataframe$'Full-time equivalent of employees, baseline year' <- as.numeric(dataframe$'Full-time equivalent of employees, baseline year')
dataframe$'Full-time equivalent of students enrolled in exclusively in distance education, performance year' <- as.numeric(dataframe$'Full-time equivalent of students enrolled in exclusively in distance education, performance year')
dataframe$'Full-time equivalent of students enrolled exclusively in distance education, baseline year' <- as.numeric(dataframe$'Full-time equivalent of students enrolled exclusively in distance education, baseline year')
dataframe$'Weighted campus users, performance year' <- as.numeric(dataframe$'Weighted campus users, performance year')
dataframe$'Weighted campus users, baseline year' <- as.numeric(dataframe$'Weighted campus users, baseline year')
dataframe$'Total waste generated per weighted campus user, performance year' <- as.numeric(dataframe$'Total waste generated per weighted campus user, performance year')
dataframe$'Total waste generated per weighted campus user, baseline year' <- as.numeric(dataframe$'Total waste generated per weighted campus user, baseline year')
dataframe$'Weighted campus users, performance year' <- as.numeric(dataframe$'Weighted campus users, performance year')
dataframe$'Weighted campus users, baseline year' <- as.numeric(dataframe$'Weighted campus users, baseline year')
dataframe$'Percentage reduction in total waste generated per weighted campus user from baseline' <- as.numeric(dataframe$'Percentage reduction in total waste generated per weighted campus user from baseline')
dataframe$'Percentage of materials diverted from the landfill or incinerator by recycling, composting, ...' <- as.numeric(dataframe$'Percentage of materials diverted from the landfill or incinerator by recycling, composting, ...')
dataframe$'Percentage of materials diverted from the landfill or incinerator (including up to 10 percent ...' <- as.numeric(dataframe$'Percentage of materials diverted from the landfill or incinerator (including up to 10 percent ...')
# Replace the single NA with 0
# Replace NA values with 0 for the specified columns
dataframe$`Total waste generated, baseline year`[is.na(dataframe$`Total waste generated, baseline year`)] <- 0
dataframe$`Weighted campus users, baseline year`[is.na(dataframe$`Weighted campus users, baseline year`)] <- 0
dataframe$`Total waste generated, performance year`[is.na(dataframe$`Total waste generated, performance year`)] <- 0
dataframe$`Weighted campus users, performance year`[is.na(dataframe$`Weighted campus users, performance year`)] <- 0
dataframe$`Materials recycled, performance year`[is.na(dataframe$`Materials recycled, performance year`)] <- 0
dataframe$`Materials composted, performance year`[is.na(dataframe$`Materials composted, performance year`)] <- 0
dataframe$`Materials donated or re-sold, performance year`[is.na(dataframe$`Materials donated or re-sold, performance year`)] <- 0
dataframe$`Materials disposed in a solid waste landfill or incinerator, performance year`[is.na(dataframe$`Materials disposed in a solid waste landfill or incinerator, performance year`)] <- 0
```

```{r}
pop_data$Institution[pop_data$Institution == "University of Minnesota Twin Cities"] <- "University of Minnesota, Twin Cities" 
pop_data$Institution[pop_data$Institution == "University of Minnesota Crookston"] <- "University of Minnesota, Crookston" 
pop_data$Institution[pop_data$Institution == "University of Minnesota Morris"] <- "University of Minnesota, Morris"
```

```{r}
# Converting 'chr' to 'date'  
dataframe$'Last Updated' <- as.Date(dataframe$'Last Updated') 
dataframe$'Start date, performance year or 3-year period' <- as.Date(dataframe$'Start date, performance year or 3-year period', origin = "1899-12-30") 
dataframe$'End date, performance year or 3-year period' <- as.Date(dataframe$'End date, performance year or 3-year period', origin = "1899-12-30") 
dataframe$'Start date, baseline year or 3-year period' <- as.Date(dataframe$'Start date, baseline year or 3-year period', origin = "1899-12-30") 
dataframe$'End date, baseline year or 3-year period' <- as.Date(dataframe$'End date, baseline year or 3-year period', origin = "1899-12-30") 
pop_data$Undergraduate.Population <- as.numeric(pop_data$Undergraduate.Population)
```

```{r}
peer_schools <- c("Iowa State University", "University of Minnesota, Twin Cities", "University of St. Thomas", 
                  "Macalester College", "Gustavus Adolphus College", "University of Wisconsin-Madison", 
                  "Marquette University", "University of Wisconsin-Eau Claire", 
                  "University of Wisconsin-Green Bay", "University of Wisconsin-Oshkosh", 
                  "University of Wisconsin-Platteville", "University of Wisconsin-Stevens Point", 
                  "University of Wisconsin-River Falls", "University of Minnesota, Morris", 
                  "University of Minnesota, Crookston","Concordia University",
                  "Carleton College", "South Dakota State University", 
                  "University of South Dakota")
schools_to_exclude <- c("Marquette University", "University of Wisconsin-Green Bay", "University of Minnesota, Crookston")
peers <-  dataframe|>
  filter(Institution %in% peer_schools)|>
  left_join(pop_data,by = "Institution")|>
  filter(Undergraduate.Population<11000)|>
  filter(!Institution %in% schools_to_exclude)|>
  mutate(
  `Reduction in Total Waste per Person` = 5 * (
    (
      (`Total waste generated, baseline year` / `Weighted campus users, baseline year`) - 
      (`Total waste generated, performance year` / `Weighted campus users, performance year`)
    ) / 
    (`Total waste generated, baseline year` / `Weighted campus users, baseline year`)
  ),
  
  `Total waste per person` = 2.78 * (
    (
      0.46 - 
      (`Total waste generated, performance year` / `Weighted campus users, performance year`) / 0.46
    )
  ),
  
  `Waste diverted from the landfill or incinerator` = 3 * (
    (
      (
        `Materials recycled, performance year` + 
        `Materials composted, performance year` + 
        `Materials donated or re-sold, performance year` +
        (0.1 * (
          `Materials recycled, performance year` + 
          `Materials composted, performance year` + 
          `Materials donated or re-sold, performance year` + 
          `Materials disposed in a solid waste landfill or incinerator, performance year`
        ))
      ) / 
      (
        `Materials recycled, performance year` + 
        `Materials composted, performance year` + 
        `Materials donated or re-sold, performance year` + 
        `Materials disposed in a solid waste landfill or incinerator, performance year`
      )
    )
  )
)
```

### Viz 1

```{r}
# Filter relevant columns for visualization
filtered_data_viz3 <- peers |>
  select(
    Institution,
    `Materials recycled, performance year`,
    `Materials composted, performance year`,
    `Materials donated or re-sold, performance year`,
    `Materials disposed in a solid waste landfill or incinerator, performance year`
  )

# Rename columns for simplicity
colnames(filtered_data_viz3) <- c(
  'Institution', 
  'Recycled', 
  'Composted', 
  'Donated or Resold', 
  'Disposed'
)

# pivot the dataframe to long format for visualization
pivot_data <- filtered_data_viz3 |>
  pivot_longer(
    cols = c('Recycled', 'Composted', 'Donated or Resold', 'Disposed'),
    names_to = 'Material_Type',
    values_to = 'Amount'
  )

ggplot(pivot_data, aes(x = Institution, y = Amount, fill = Material_Type)) +
  geom_bar(stat = 'identity') +
  scale_fill_manual(
    values = c(
      'Recycled' = 'blue',
      'Composted' = 'darkgreen',
      'Donated or Resold' = 'orange',
      'Disposed' = 'red'
    )
  ) +
  labs(
    x = 'Institution',
    y = 'Total Amount of Material (tons)',
    title = 'Material Disposal Methods by Institution',
    fill = 'Material Type'
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))


```

The University of St. Thomas recycles significantly more materials than the overall average of other schools. However, UST composts less compared to the overall average and donates/resells fewer materials relative to the average. Notably, UST disposes of less material in landfills or incinerators compared to the overall average, indicating stronger waste diversion efforts.

### Viz 2

```{r}
filtered_data_viz4 <- peers|>
  select(Institution, Undergraduate.Population, 'Percentage reduction in total waste generated per weighted campus user from baseline') |>
  rename(
    School = Institution,
    Population = Undergraduate.Population,
    Total_Waste_Reduction = 'Percentage reduction in total waste generated per weighted campus user from baseline'
  ) |>
  filter(!is.na(School) & !is.na(Population) & !is.na(Total_Waste_Reduction))

# Create a column to distinguish University of St. Thomas from other schools
filtered_data_viz4 <- filtered_data_viz4 |>
  mutate(Color = ifelse(grepl('University of St. Thomas', School), 'purple', 'grey'))

# Create the scatter plot
ggplot(filtered_data_viz4, aes(x = School, y = Total_Waste_Reduction, size = Population, color = Color)) +
  geom_point(shape = 21, fill = filtered_data_viz4$Color) +  # Change shape to circle (shape = 21) or square (shape = 22)
  scale_color_manual(values = c('purple' = 'purple', 'grey' = 'grey')) +
  scale_size(range = c(3, 10)) +  # Adjust the size of the points
  labs(
    x = 'School',
    y = 'Total Reduction in Waste',
    title = 'Comparison of Schools by Total Waste Reduction Percentage'
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1),
    legend.position="none"
  )


```

Compared to other institutions, the University of St. Thomas' total reduction in waste is relatively higher and its population size appears to be similarly relative to the majority of other schools. This suggests that St. Thomas's waste reduction efforts are more effective based on the data when considering its population size.
