---
title: "PHASE 2"
author: "Nimo Hassan, Alexander Wodarski, Breanna Ranglall, Tenzin Dhondup"
format: pdf
editor: visual
---

```{r}
#| warning: false
library(tidyverse)
library(readxl)
library(dplyr)
```

```{r}
# Read data from excel spreadsheet
dataf <- read_excel("OP-18_Waste_Minimization_and_Diversion_2024.xlsx")
pop_data <- read.csv("peerspop.csv")
```

```{r}
#| warning: false
dataframe <- dataf

# Converting 'chr' to 'numeric'

dataframe$'Materials recycled, performance year' <- as.numeric(dataframe$'Materials recycled, performance year')
dataframe$'Materials recycled, baseline year' <- as.numeric(dataframe$'Materials recycled, baseline year')
dataframe$'Materials composted, performance year' <- as.numeric(dataframe$'Materials composted, performance year')
dataframe$'Materials composted, baseline year' <- as.numeric(dataframe$'Materials composted, baseline year')
dataframe$'Materials donated or re-sold, performance year' <- as.numeric(dataframe$'Materials donated or re-sold, performance year')
dataframe$'Materials donated or re-sold, baseline year' <- as.numeric(dataframe$'Materials donated or re-sold, baseline year')
dataframe$'Materials disposed through post-recycling residual conversion, performance year' <- as.numeric(dataframe$'Materials disposed through post-recycling residual conversion, performance year')
dataframe$'Materials disposed through post-recycling residual conversion, baseline year' <- as.numeric(dataframe$'Materials disposed through post-recycling residual conversion, baseline year')
dataframe$'Materials disposed in a solid waste landfill or incinerator, performance year' <- as.numeric(dataframe$'Materials disposed in a solid waste landfill or incinerator, performance year')
dataframe$'Materials disposed in a solid waste landfill or incinerator, baseline year' <- as.numeric(dataframe$'Materials disposed in a solid waste landfill or incinerator, baseline year')
dataframe$'Total waste generated, performance year' <- as.numeric(dataframe$'Total waste generated, performance year')
dataframe$'Total waste generated, baseline year' <- as.numeric(dataframe$'Total waste generated, baseline year')
dataframe$'Start date, performance year or 3-year period' <- as.numeric(dataframe$'Start date, performance year or 3-year period')
dataframe$'End date, performance year or 3-year period' <- as.numeric(dataframe$'End date, performance year or 3-year period')
dataframe$'Start date, baseline year or 3-year period' <- as.numeric(dataframe$'Start date, baseline year or 3-year period')
dataframe$'End date, baseline year or 3-year period' <- as.numeric(dataframe$'End date, baseline year or 3-year period')
dataframe$'Number of students resident on-site, performance year' <- as.numeric(dataframe$'Number of students resident on-site, performance year')
dataframe$'Number of students resident on-site, baseline year' <- as.numeric(dataframe$'Number of students resident on-site, baseline year')
dataframe$'Number of employees resident on-site, performance year' <- as.numeric(dataframe$'Number of employees resident on-site, performance year')
dataframe$'Number of employees resident on-site, baseline year' <- as.numeric(dataframe$'Number of employees resident on-site, baseline year')
dataframe$'Number of other individuals resident on-site, performance year' <- as.numeric(dataframe$'Number of other individuals resident on-site, performance year')
dataframe$'Number of other individuals resident on-site, baseline year' <- as.numeric(dataframe$'Number of other individuals resident on-site, baseline year')
dataframe$'Total full-time equivalent student enrollment, performance year' <- as.numeric(dataframe$'Total full-time equivalent student enrollment, performance year')
dataframe$'Total full-time equivalent student enrollment, baseline year' <- as.numeric(dataframe$'Total full-time equivalent student enrollment, baseline year')
dataframe$'Full-time equivalent of employees, performance year' <- as.numeric(dataframe$'Full-time equivalent of employees, performance year')
dataframe$'Full-time equivalent of employees, baseline year' <- as.numeric(dataframe$'Full-time equivalent of employees, baseline year')
dataframe$'Full-time equivalent of students enrolled in exclusively in distance education, performance year' <- as.numeric(dataframe$'Full-time equivalent of students enrolled in exclusively in distance education, performance year')
dataframe$'Full-time equivalent of students enrolled exclusively in distance education, baseline year' <- as.numeric(dataframe$'Full-time equivalent of students enrolled exclusively in distance education, baseline year')
dataframe$'Weighted campus users, performance year' <- as.numeric(dataframe$'Weighted campus users, performance year')
dataframe$'Weighted campus users, baseline year' <- as.numeric(dataframe$'Weighted campus users, baseline year')
dataframe$'Total waste generated per weighted campus user, performance year' <- as.numeric(dataframe$'Total waste generated per weighted campus user, performance year')
dataframe$'Total waste generated per weighted campus user, baseline year' <- as.numeric(dataframe$'Total waste generated per weighted campus user, baseline year')
dataframe$'Weighted campus users, performance year' <- as.numeric(dataframe$'Weighted campus users, performance year')
dataframe$'Weighted campus users, baseline year' <- as.numeric(dataframe$'Weighted campus users, baseline year')
dataframe$'Percentage reduction in total waste generated per weighted campus user from baseline' <- as.numeric(dataframe$'Percentage reduction in total waste generated per weighted campus user from baseline')
dataframe$'Percentage of materials diverted from the landfill or incinerator by recycling, composting, ...' <- as.numeric(dataframe$'Percentage of materials diverted from the landfill or incinerator by recycling, composting, ...')
dataframe$'Percentage of materials diverted from the landfill or incinerator (including up to 10 percent ...' <- as.numeric(dataframe$'Percentage of materials diverted from the landfill or incinerator (including up to 10 percent ...')
# Replace the single NA with 0
dataframe$`Materials donated or re-sold, performance year`[is.na(dataframe$`Materials donated or re-sold, performance year`)] <- 0

```

```{r}
# Converting 'chr' to 'date'

dataframe$'Last Updated' <- as.Date(dataframe$'Last Updated')
dataframe$'Start date, performance year or 3-year period' <- as.Date(dataframe$'Start date, performance year or 3-year period', origin = "1899-12-30")
dataframe$'End date, performance year or 3-year period' <- as.Date(dataframe$'End date, performance year or 3-year period', origin = "1899-12-30")
dataframe$'Start date, baseline year or 3-year period' <- as.Date(dataframe$'Start date, baseline year or 3-year period', origin = "1899-12-30")
dataframe$'End date, baseline year or 3-year period' <- as.Date(dataframe$'End date, baseline year or 3-year period', origin = "1899-12-30")
pop_data$Undergraduate.Population <- as.numeric(pop_data$Undergraduate.Population)
```

```{r}
peer_schools<-c("Bemidji State University","Carleton College","College of St. Benedict/St. John’s University", "Concordia Moorhead","Macalester College","Winona State University","University of Minnesota, Twin Cities","University of Minnesota, Morris","UMN – Duluth","Augsburg University","Concordia in St. Paul","Hamline University","St. Kate’s University","St. Olaf College","University of St. Thomas") 
peers <-  dataframe|>
  filter(Institution %in% peer_schools)|>
  left_join(pop_data,by = "Institution")|>
  mutate(
  `Reduction in Total Waste per Person` = 5 * (
    (
      (`Total waste generated, baseline year` / `Weighted campus users, baseline year`) - 
      (`Total waste generated, performance year` / `Weighted campus users, performance year`)
    ) / 
    (`Total waste generated, baseline year` / `Weighted campus users, baseline year`)
  ),
  
  `Total waste per person` = 2.78 * (
    (
      0.46 - 
      (`Total waste generated, performance year` / `Weighted campus users, performance year`) / 0.46
    )
  ),
  
  `Waste diverted from the landfill or incinerator` = 3 * (
    (
      (
        `Materials recycled, performance year` + 
        `Materials composted, performance year` + 
        `Materials donated or re-sold, performance year` +
        (0.1 * (
          `Materials recycled, performance year` + 
          `Materials composted, performance year` + 
          `Materials donated or re-sold, performance year` + 
          `Materials disposed in a solid waste landfill or incinerator, performance year`
        ))
      ) / 
      (
        `Materials recycled, performance year` + 
        `Materials composted, performance year` + 
        `Materials donated or re-sold, performance year` + 
        `Materials disposed in a solid waste landfill or incinerator, performance year`
      )
    )
  )
)

           
# Summary statistics for all numeric variables
numeric_summary <- peers |>
  summarize(across(where(is.numeric), list(
    mean = ~ mean(.x, na.rm = TRUE),
    median = ~ median(.x, na.rm = TRUE),
    sd = ~ sd(.x, na.rm = TRUE),
    min = ~ min(.x, na.rm = TRUE),
    max = ~ max(.x, na.rm = TRUE)
  )))




View(peers)
View(numeric_summary) 
print(numeric_summary) 

```

We can compare St. Thomas with all the summary statistics. One thing to note out of the peer institutions, is that the University of Minnesota Twin Cities skews the results as it has much larger numbers than all the other Institutions that we are comparing too. Because of this St. Thomas ends up being below the average for the statistics. The main assumption for the reason would be that the University of Minnesota Twin Cities is just a much larger school compared to the others on the list and therefore their numbers are much higher. St Thomas rank fluctuates as 2 or 3 for the variables behind the University of Minnesota, and they fluctuate with Carleton College whon has similar numbers for most.

```{r}
library(dplyr)


summarydata <- data.frame(
  
  Material = rep(c(
    "Materials recycled",
    "Materials composted",
    "Materials donated or re-sold",
    "Materials disposed through post-recycling residual conversion",
    "Materials disposed in a solid waste landfill or incinerator"
  ), each = 10),
  Year_Type = rep(c(
    "Performance Year", "Baseline Year"
  ), times = 5, each = 5),
  Metric = rep(c("Mean", "Median", "SD", "Min", "Max"), times = 10),
  Value = c(
    451.446, 229.74, 538.0773, 105, 1390.34,
    675.482, 525, 869.5525, 57, 2176.4,
    308.51, 67.4, 488.207, 53.05, 1176.1,
    215.175, 215.175, 208.7733, 67.55, 362.8,
    68.52, 6.6, 136.8534, 0, 312.84,
    107.52, 107.52, 63.32848, 62.74, 152.3,
    NA, NA, NA, Inf, -Inf,
    NA, NA, NA, Inf, -Inf,
    934.404, 358.09, 1558.256, 104, 3712.9,
    934.404, 358.09, 1558.256, 104, 3712.9
  )
  
)


organizedSummary <- summarydata|>
  pivot_wider(
    names_from = Metric,
    values_from = Value
  ) |>
  arrange(Material, Year_Type)


view(organizedSummary)

```

### Pull in Clean Data

```{r}
# Read in the CSV file
df_pop <- read.csv("peerspop.csv")

# Clean and prepare the data
clean_df_pop <- df_pop |>
  select(
    Institution, 
    Materials.recycled..performance.year, 
    Materials.composted..performance.year, 
    Undergraduate.Population
  ) |>
  mutate(
    Materials.recycled = as.numeric(Materials.recycled..performance.year),
    Materials.composted = as.numeric(Materials.composted..performance.year),
    Undergraduate.Population = as.numeric(Undergraduate.Population)
  ) |>
  filter(!is.na(Materials.recycled), !is.na(Materials.composted), !is.na(Undergraduate.Population))

```

### Viz 1

```{r}
# Plot 1: Scatter Plot of Materials Composted vs. Materials Recycled
ggplot(clean_df_pop, aes(x = Materials.recycled, y = Materials.composted, size = Undergraduate.Population, color = Institution)) +
  geom_point(shape = 15) + # Shape changed to square
  labs(
    title = "Materials Composted vs. Materials Recycled by Institution (Bubble Size = Population)",
    x = "Materials Recycled (Tons)",
    y = "Materials Composted (Tons)",
    caption = 'This line plot compares the recycled and composted materials for each institution. \nThe University of St. Thomas can be observed relative to other universities with similar \npopulation, allowing for insights into its recycling and composting efforts.'

  ) +
  scale_size_continuous(range = c(3, 15)) +
  theme_minimal()

```

### Viz 2

```{r}
# Plot 2: Line Plot of Recycled and Composted Materials for Institutions

# Arrange data to ensure institutions are in order
clean_df_pop_sorted <- clean_df_pop |>
  arrange(Institution)

# Create the line plot
ggplot(clean_df_pop_sorted, aes(x = Institution)) +
  geom_line(aes(y = `Materials.recycled`, color = 'Recycled Materials'), group = 1) +
  geom_point(aes(y = `Materials.recycled`, color = 'Recycled Materials')) +
  geom_line(aes(y = `Materials.composted`, color = 'Composted Materials'), group = 1) +
  geom_point(aes(y = `Materials.composted`, color = 'Composted Materials')) +
  scale_color_manual(values = c('Recycled Materials' = 'blue', 'Composted Materials' = 'green')) +
  labs(
    title = 'Recycled and Composted Materials by Institution (Performance Year)',
    x = 'Institution',
    y = 'Amount (Tons)',
    color = 'Material Type',
    caption = 'This line plot compares the recycled and composted materials for each institution. \nThe University of St. Thomas can be observed relative to other universities, \nallowing for insights into its recycling and composting efforts.'

  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

### Viz 3

```{r}
# Converting df from wide format to long format

df_long <- dataframe |>
  select(Institution, where(is.numeric)) |>
  pivot_longer(
    cols = starts_with("Materials") | starts_with("Total") | starts_with("Number") | starts_with("Full-Time") | starts_with("Weighted"),
    names_to = c("Variable_Type", "Year"),
    names_sep = ", ",
    values_to = "Amount"
  )
```

```{r}
#Combining Tidy Numeric Data with Non Numeric Data

df_non_numeric <- dataframe |>
  select(where(~ !is.numeric(.)))

data_test <- left_join(df_long, df_non_numeric, by = "Institution")
```

```{r}
# Filter waste and enrollment data separately, then join them
waste_data <- data_test |>
  filter(Institution %in% c("University of St. Thomas", "Carleton College", "Concordia University", "Macalester College", "University of Minnesota, Twin Cities", "University of Minnesota, Morris", "University of Minnesota, Duluth")) |>
  filter(Variable_Type == "Total waste generated") |>
  filter(Year == "performance year") |>
  filter(!is.na(Amount)) |>
  select(Institution, Amount, Variable_Type, Year)

enrollment_data <- data_test |>
  filter(Institution %in% c("University of St. Thomas", "Carleton College", "Concordia University", "Macalester College", "University of Minnesota, Twin Cities", "University of Minnesota, Morris", "University of Minnesota, Duluth")) |>
  filter(Variable_Type == "Total full-time equivalent student enrollment") |>
  filter(Year == "performance year") |>
  filter(!is.na(Amount)) |>
  select(Institution, Amount, Variable_Type, Year)

# Join the two datasets on Institution
merged_data <- left_join(waste_data, enrollment_data, by = "Institution", suffix = c("_waste", "_enrollment"))

# Plot the data with a legend
ggplot(merged_data, aes(x = Amount_waste, y = Amount_enrollment)) +
  geom_point(aes(color = Institution), size = 3) +  # Mapping color to Institution for legend
  scale_color_manual(values = c("University of St. Thomas" = "purple", 
                                "Carleton College" = "grey40", 
                                "Concordia University" = "grey60", 
                                "Macalester College" = "grey80", 
                                "University of Minnesota, Twin Cities" = "grey30", 
                                "University of Minnesota, Morris" = "grey50", 
                                "University of Minnesota, Duluth" = "grey70")) +  # Set grey shades manually
  labs(title = "Waste Generation vs. Campus Population",
       x = "Total Waste Generated (Performance Year)",
       y = "Full-Time Equivalent Students (Performance Year)",
       color = "Institution") +  # Legend title
  theme_minimal()
```

### Viz 4

```{r}
# Bar Plot: Total Waste Generated by Institution
ggplot(merged_data, aes(x = Institution, y = Amount_waste, fill = Institution)) +
  geom_bar(stat = "identity", show.legend = FALSE) +
  scale_fill_manual(values = c("University of St. Thomas" = "purple", 
                                "Carleton College" = "grey40", 
                                "Concordia University" = "grey60", 
                                "Macalester College" = "grey80", 
                                "University of Minnesota, Twin Cities" = "grey30", 
                                "University of Minnesota, Morris" = "grey50", 
                                "University of Minnesota, Duluth" = "grey70")) +  # Set grey shades manually
  labs(title = "Total Waste Generated by Institution",
       x = "Institution",
       y = "Total Waste Generated (Performance Year)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```
